---
title: "Reporting disease frequency in clinical research: incidence and prevalence"
authors: "Renato Carneiro de Freitas Chaves, Tiago Mendonça dos Santos, Thiago Domingos Corrêa"
date: "`r Sys.Date()`"
output: html_document
    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# 1) Purpose

This R Markdown document provides a step-by-step analytical guide for estimating and interpreting core epidemiological measures of disease frequency using the attached clinical dataset.

The analysis focuses on four related concepts:

1. **Incidence proportion**: the proportion of initially at-risk patients who experienced the event during the observation period.
2. **Incidence rate**: the number of new events divided by the amount of person-time contributed by the at-risk population.
3. **Prevalence**: the proportion of patients with a condition at a defined time point, during a defined period, or among individuals who had contact with the health-care setting.
4. **At-risk population**: the group of individuals who are eligible to experience the outcome of interest.

The dataset contains individual-level clinical information for patients admitted to the intensive care unit. In this didactic example, hospital mortality at discharge is represented by outcome = 1. Follow-up-based events are represented by event = 1 and are used for person-time analyses.

the main incident event is **hospital mortality** (`outcome = 1`). 
> **Dataset coding according to the legend**  
> - `group`: `0 = Control`, `1 = Intervention`  
> - `outcome`: `0 = Alive`, `1 = Dead`  
> - Binary comorbidity and support variables: `0 = No`, `1 = Yes`  
> - `mechanical_ventilation`: use of invasive mechanical ventilation at ICU admission  
> - `vasopressor`: vasopressor use at ICU admission  

Note
> -  'outcome': status at hospital discharge (time-independent)
> - 'event': occurrence during follow-up (time-dependent, used for incidence rate)

# 2) Packages

```{r packages}
# Install packages once if necessary:
# install.packages(c("readxl", "dplyr", "janitor", "tibble", "knitr", "skimr", "lubridate"))

library(readxl)
library(dplyr)
library(janitor)
library(tibble)
library(knitr)
library(skimr)
library(lubridate)
```

# 3) Data Import (Excel)
This guide uses the Excel file **data.xlsx** with a sheet named **"Data"**.

- If `data.xlsx` is in the same folder as this `.Rmd`, you can keep the path as `"data.xlsx"`.
- Otherwise, replace with the full path to your file.

```{r data-loading}
data <- read_excel("data.xlsx", sheet = "Data") 

```

# 4) Confirm Required Variables

For the requested calculations, we need:
- `group` (0 = Control, 1 = Intervention)
- `outcome` (0 = Alive, 1 = Dead)
- `event` (0 = censored/no event during follow-up, 1 = event occurred)
- `follow_up_start` (start date of follow-up)
- `follow_up_end` (end date of follow-up)
- `follow_up_days` (duration of follow-up in days)
- `follow_up_years` (duration of follow-up in years)

```{r verify-vars-2}
# Quick frequency checks
skimr::skim(data)

data %>% 
  tabyl(outcome)

data %>% 
  tabyl(event)

summary(data$follow_up_days)
summary(data$follow_up_years)

```

# 5) Data Preparation and Variable Definitions

The analysis uses the following variables:

- `patient`: unique patient identifier.
- `group`: patient allocation group.
- `follow_up_start`: start date of follow-up.
- `follow_up_end`: end date of follow-up, corresponding either to event occurrence or censoring.
- `follow_up_days`: individual follow-up duration in days.
- `follow_up_years`: individual follow-up duration in years, calculated as `follow_up_days / 365.25`.
- `outcome`: hospital vital status; `1 = death`, `0 = alive`.
- `event`: event indicator during follow-up; `1 = event occurred`, `0 = censored/no event observed`.
- `mechanical_ventilation`: invasive mechanical ventilation at ICU admission.
- `vasopressor`: vasopressor use at ICU admission.
- `comorbidity_any`: presence of at least one selected chronic comorbidity.
- `person_time_days`: amount of observation time contributed by each at-risk patient in days.
- `person_time_years`: amount of observation time contributed by each at-risk patient in years.

```{r data-preparation}
analysis_data <- data %>%
  # Keep variables required for this epidemiological analysis.
  select(
    patient, group, center,
    follow_up_start, follow_up_end, follow_up_days, follow_up_years,
    age, platelet, hemoglobin, creatinine, saps_3, fluid_balance,
    hypertension, diabetes, heart_failure, smoking,
    chronic_kidney_disease, cancer, stroke,
    vasopressor, mechanical_ventilation,
    outcome, event
  ) %>%
  # Convert coded variables and follow-up fields to the appropriate analytical format.
  mutate(
    across(
      c(group, center, outcome, event, hypertension, diabetes, heart_failure,
        smoking, chronic_kidney_disease, cancer, stroke,
        vasopressor, mechanical_ventilation),
      as.integer
    ),

    # Convert follow-up dates to Date objects.
    # This ensures that date subtraction and time calculations are reproducible.
    follow_up_start = as.Date(follow_up_start),
    follow_up_end   = as.Date(follow_up_end),

    # Validate follow-up duration.
    # The Excel file already contains follow_up_days and follow_up_years.
    # Here, we recalculate follow-up duration from dates only as a consistency check.
    calculated_follow_up_days = as.numeric(follow_up_end - follow_up_start),

    # Use the follow-up duration stored in the dataset for the primary analysis.
    # In a real cohort, this value would represent time from cohort entry to event or censoring.
    person_time_days = as.numeric(follow_up_days),
    person_time_years = as.numeric(follow_up_years),

    # Define the at-risk indicator.
    # Patients are considered at risk when they have valid event status and positive follow-up time.
    at_risk_discharge = as.integer(!is.na(outcome)),
    
    at_risk_followup = as.integer(
      !is.na(event) &
        !is.na(person_time_years) &
        person_time_years > 0
),
    # Define a composite chronic comorbidity variable.
    # This represents whether at least one selected chronic condition is documented.
    comorbidity_any = as.integer(
      hypertension == 1 | diabetes == 1 | heart_failure == 1 |
        smoking == 1 | chronic_kidney_disease == 1 | cancer == 1 | stroke == 1
    ),

    # Create readable labels for descriptive tables.
    group_label = factor(group, levels = c(0, 1), labels = c("Control", "Intervention")),
    outcome_label = factor(outcome, levels = c(0, 1), labels = c("Alive", "Dead")),
    event_label = factor(event, levels = c(0, 1), labels = c("Censored", "Event occurred"))
  )

# Confirm that the prepared dataset has the expected analytical variables.
glimpse(analysis_data)


```

# 6) Describing Disease Frequency

1. **The numerator**: number of patients with the disease or event.
2. **The denominator**: population from which the numerator arose.
3. **The time frame**: point in time, period of observation, or accumulated person-time.
4. **The population context**: inclusion criteria and at-risk status.


```{r descriptive-frequency}
# Distribution of the main event by allocation group.
event_by_group <- analysis_data %>%
  count(group_label, outcome_label, name = "n") %>%
  group_by(group_label) %>%
  mutate(group_proportion = n / sum(n)) %>%
  ungroup()

event_by_group %>%
  kable(
    digits = 3,
    caption = "Hospital vital status by allocation group."
  )
```

# 7) The At-Risk Population

## 7.1 Conceptual definition

The at-risk population includes individuals who are eligible to experience the event of interest. For hospital mortality, patients must be alive at the beginning of follow-up and have a known hospital outcome.

## 7.2 Mathematical notation

Let:

\[N_{risk} = \sum_{i=1}^{N} I(\text{patient } i \text{ is at risk})\]

where \(I(\cdot)\) is an indicator function equal to 1 when the condition is true and 0 otherwise.

```{r at-risk-population}
# Count the at-risk population for the main event.
discharge_risk_set <- analysis_data %>%
  filter(at_risk_discharge == 1)

incidence_proportion_overall <- discharge_risk_set %>%
  summarise(
    population_at_risk = n(),
    incident_events = sum(outcome == 1, na.rm = TRUE),
    incidence_proportion = incident_events / population_at_risk,
    incidence_percent = incidence_proportion * 100
  )

```

# 8) Incidence Proportion

## 8.1 Definition

Incidence proportion, also called cumulative incidence or risk, is the proportion of at-risk individuals who experience a new event during a defined observation period.

## 8.2 Formula

\[\text{Incidence proportion} =
\frac{\text{Number of new events during the period}}
{\text{Number of individuals at risk at the beginning of the period}}
\]

For percentage reporting:

\[
\text{Incidence proportion (\%)} =
\text{Incidence proportion} \times 100
\]

```{r incidence-proportion}

# Calculate incidence proportion
discharge_risk_set <- analysis_data %>%
  filter(at_risk_discharge == 1)

incidence_proportion_overall <- discharge_risk_set %>%
  summarise(
    population_at_risk = n(),
    incident_events = sum(outcome == 1, na.rm = TRUE),
    incidence_proportion = incident_events / population_at_risk,
    incidence_percent = incidence_proportion * 100
  )
```

# 9) Incidence Rate

## 9.1 Definition

Incidence rate is the number of new events divided by the total person-time accumulated by the at-risk population.

## 9.2 Formula

First, total person-time is calculated as the sum of individual follow-up time:

\[
\text{Total person-time} = \sum_{i=1}^{N} t_i
\]

where \(t_i\) is the follow-up time contributed by individual \(i\).

Then, the incidence rate is calculated as:

\[
\text{Incidence rate} =
\frac{\text{Number of new events}}
{\text{Total person-time at risk}}
\]

For standardized reporting:

\[
\text{Incidence rate per 100 person-years} =
\text{Incidence rate} \times 100
\]

\[
\text{Incidence rate per 1,000 person-years} =
\text{Incidence rate} \times 1000
\]


```{r incidence-rate}
# Calculate total person-time and the event rate for the full cohort.
# The denominator is the sum of each patient's follow-up time.
followup_risk_set <- analysis_data %>%
  filter(
    at_risk_followup == 1
  )

incidence_rate_overall <- followup_risk_set %>%
  summarise(
    population_at_risk = n(),
    incident_events = sum(event == 1, na.rm = TRUE),
    total_person_time_days = sum(person_time_days, na.rm = TRUE),
    total_person_time_years = sum(person_time_years, na.rm = TRUE),
    incidence_rate_per_person_year = incident_events / total_person_time_years,
    incidence_rate_per_100_person_years = incidence_rate_per_person_year * 100,
    incidence_rate_per_1000_person_years = incidence_rate_per_person_year * 1000
  )


# Calculate incidence rate by allocation group.
incidence_rate_by_group <- followup_risk_set %>%
  group_by(group_label) %>%
  summarise(
    population_at_risk = n(),
    incident_events = sum(event == 1, na.rm = TRUE),
    total_person_time_days = sum(person_time_days, na.rm = TRUE),
    total_person_time_years = sum(person_time_years, na.rm = TRUE),
    incidence_rate_per_person_year = incident_events / total_person_time_years,
    incidence_rate_per_100_person_years = incidence_rate_per_person_year * 100,
    incidence_rate_per_1000_person_years = incidence_rate_per_person_year * 1000,
    .groups = "drop"
  )


```

# 10) Prevalence

Prevalence measures the proportion of a population that has a condition at a defined point or during a defined interval. 

## 10.1 Point prevalence

### Definition

Point prevalence is the proportion of individuals with a condition at a specific point in time. In this dataset, invasive mechanical ventilation is documented at ICU admission, so it can be interpreted as a point-prevalence measure at the time of ICU admission.

### Formula

\[
\text{Point prevalence} =
\frac{\text{Number of individuals with the condition at a specific point}}
{\text{Number of individuals assessed at that point}}
\]

```{r point-prevalence}
# Calculate point prevalence of invasive mechanical ventilation at ICU admission.
point_prevalence_mv <- analysis_data %>%
  filter(!is.na(mechanical_ventilation)) %>%
  summarise(
    assessed_n = n(),
    prevalent_cases = sum(mechanical_ventilation == 1, na.rm = TRUE),
    point_prevalence = prevalent_cases / assessed_n,
    point_prevalence_percent = point_prevalence * 100
  )

```

## 10.2 Period prevalence

### Definition

Period prevalence is the proportion of individuals who had a condition at any time during a defined observation period. In this dataset, chronic comorbidities are documented for patients enrolled during the study period. The composite variable `comorbidity_any` is used as a period-prevalence example.

### Formula

\[
\text{Period prevalence} =
\frac{\text{Number of individuals with the condition at any time during the period}}
{\text{Number of individuals observed during the same period}}
\]

```{r period-prevalence}
# Calculate period prevalence of at least one selected chronic comorbidity.
period_prevalence_comorbidity <- analysis_data %>%
  filter(!is.na(comorbidity_any)) %>%
  summarise(
    observed_n = n(),
    prevalent_cases = sum(comorbidity_any == 1, na.rm = TRUE),
    period_prevalence = prevalent_cases / observed_n,
    period_prevalence_percent = period_prevalence * 100
  )

```

## 10.3 Contact prevalence

### Definition

Contact prevalence is the proportion of individuals with a condition among those who had contact with a defined health-care service or setting. Here, all enrolled patients had contact with the ICU study setting. Therefore, contact prevalence is illustrated as the prevalence of vasopressor use among ICU contacts.

### Formula

\[
\text{Contact prevalence} =
\frac{\text{Number of contacts with the condition}}
{\text{Total number of contacts assessed}}
\]

```{r contact-prevalence}
# Calculate contact prevalence of vasopressor use among enrolled ICU contacts.
contact_prevalence_vasopressor <- analysis_data %>%
  filter(!is.na(vasopressor)) %>%
  summarise(
    contacts_assessed = n(),
    contacts_with_condition = sum(vasopressor == 1, na.rm = TRUE),
    contact_prevalence = contacts_with_condition / contacts_assessed,
    contact_prevalence_percent = contact_prevalence * 100
  )

```

# 11) Contingency Tables

Contingency tables are useful for displaying disease frequency across clinically meaningful strata.

```{r contingency-tables}
# Event by allocation group.
tab <- analysis_data %>%
  tabyl(group_label, outcome_label) %>%
  adorn_totals(where = c("row", "col"))


# Mechanical ventilation by allocation group.
table_mv_group <- analysis_data %>%
  mutate(
    mechanical_ventilation_label = factor(
      mechanical_ventilation,
      levels = c(0, 1),
      labels = c("No mechanical ventilation", "Mechanical ventilation")
    )
  ) %>%
  tabyl(group_label, mechanical_ventilation_label) %>%
  adorn_totals(where = c("row", "col"))
```

# 12) Disease Frequency Across Selected Conditions

The same prevalence structure can be applied to multiple binary clinical variables. This table summarizes the frequency of selected comorbidities and ICU-support variables.

```{r selected-condition-frequency}
selected_binary_variables <- c(
  "hypertension",
  "diabetes",
  "heart_failure",
  "smoking",
  "chronic_kidney_disease",
  "cancer",
  "stroke",
  "vasopressor",
  "mechanical_ventilation"
)

condition_frequency <- lapply(selected_binary_variables, function(variable_name) {
  x <- analysis_data[[variable_name]]

  tibble(
    variable = variable_name,
    assessed_n = sum(!is.na(x)),
    cases_n = sum(x == 1, na.rm = TRUE),
    prevalence = cases_n / assessed_n,
    prevalence_percent = prevalence * 100
  )
}) %>%
  bind_rows()

```

# 13) Final Summary Table

```{r final-summary}
final_summary_table <- bind_rows(
  tibble(
    measure = "Incidence proportion of hospital mortality at discharge",
    numerator = incidence_proportion_overall$incident_events,
    denominator = incidence_proportion_overall$population_at_risk,
    estimate = incidence_proportion_overall$incidence_percent,
    unit = "Percent"
  ),
  tibble(
    measure = "Incidence rate during follow-up",
    numerator = incidence_rate_overall$incident_events,
    denominator = incidence_rate_overall$total_person_time_years,
    estimate = incidence_rate_overall$incidence_rate_per_100_person_years,
    unit = "Events per 100 person-years"
  ),
  tibble(
    measure = "Point prevalence of invasive mechanical ventilation",
    numerator = point_prevalence_mv$prevalent_cases,
    denominator = point_prevalence_mv$assessed_n,
    estimate = point_prevalence_mv$point_prevalence_percent,
    unit = "Percent"
  ),
  tibble(
    measure = "Period prevalence of at least one selected chronic comorbidity",
    numerator = period_prevalence_comorbidity$prevalent_cases,
    denominator = period_prevalence_comorbidity$observed_n,
    estimate = period_prevalence_comorbidity$period_prevalence_percent,
    unit = "Percent with at least one selected chronic comorbidity"
  ),
  tibble(
    measure = "Contact prevalence of vasopressor use among ICU contacts",
    numerator = contact_prevalence_vasopressor$contacts_with_condition,
    denominator = contact_prevalence_vasopressor$contacts_assessed,
    estimate = contact_prevalence_vasopressor$contact_prevalence_percent,
    unit = "Percent of enrolled ICU contacts receiving vasopressors"
  )
)

final_summary_table %>%
  kable(
    digits = 0,
    caption = "Reporting disease frequency in clinical research"
  )
```


## End of document
