---

title: "Survival analysis in clinical and public health research"
authors: "Renato Carneiro de Freitas Chaves, Tiago Mendonça dos Santos, Thiago Domingos Corrêa"
date: "`r Sys.Date()`"
output: html_document

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# 1) Purpose

This R Markdown document provides a step-by-step introduction to survival analysis using a synthetic clinical dataset. The primary objective is to compare time to death between an intervention group and a control group while accounting for unequal follow-up and right censoring.

The analysis addresses five questions:

1. What is the estimated probability of remaining alive over time?
2. Do the survival functions differ between the two groups?
3. What is the intervention-to-control hazard ratio?
4. Does adjustment for a prespecified baseline severity measure materially change the estimated association?
5. Is the proportional-hazards assumption reasonable?


# 2) Packages

```{r packages}
# Install these packages once if necessary:
# install.packages(c("readxl", "survival", "knitr"))

library(readxl)   
library(survival) 
library(knitr)    
```

# 3) Data import

This guide uses the Excel file **data.xlsx** with a sheet named **"Data"**.

- If `data.xlsx` is in the same folder as this `.Rmd`, you can keep the path as `"data.xlsx"`.
- Otherwise, replace with the full path to your file.

```{r import-data}
data <- read_excel("data.xlsx", sheet = "Data") 

```
# 4)  Variable Definitions and Data Inspection
Before conducting statistical analysis, the investigator should inspect the structure of the dataset, confirm the variable names, and review the first rows.


The main variables used in this report are:
- `follow_up_days`: observed time from the defined baseline to death or censoring, measured in days.
- `event`: event indicator, coded as 1 for death during follow-up and 0 for right censoring.
- `group`: allocation group, coded as 0 for control and 1 for intervention.
- `saps_3`: Simplified Acute Physiology Score 3 at ICU admission.
- `saps_3_10`: SAPS 3 divided by 10, so its Cox-model hazard ratio corresponds to a 10-point increase.
- `outcome`: vital status at hospital discharge, coded as 0 for alive and 1 for dead. This variable is not used as the survival-event indicator because it refers to a different observation window.



# Data inspection and validation

The investigator should confirm the dimensions, variable names, coding, and first observations before fitting any model.

```{r inspect-data}
# Number of rows and columns.
dim(data)

# Variable names.
names(data)

# First six records.
head(data)
```

# 5) Basic Data Preparation
The code below checks that the required variables exist, converts them to numeric form, and verifies the survival coding.
The primary unadjusted analyses require only follow-up time, event status, and group. The adjusted model additionally requires SAPS 3. Keeping separate analytic datasets prevents missing covariates from unnecessarily excluding participants from the Kaplan-Meier, log-rank, or unadjusted Cox analyses.


```{r validate-data}
required_variables <- c("follow_up_days", "event", "group", "saps_3", "outcome")

# Convert only variables used in the analysis.
data[required_variables] <- lapply(data[required_variables], as.numeric)

survival_data <- data[complete.cases(data[, c("follow_up_days", "event", "group")]),]

adjusted_data <- data[complete.cases(data[, c("follow_up_days", "event", "group", "saps_3")]),]

survival_data$group <- factor(
  survival_data$group,
  levels = c(0, 1),
  labels = c("Control", "Intervention")
)

adjusted_data$group <- factor(
  adjusted_data$group,
  levels = c(0, 1),
  labels = c("Control", "Intervention")
)

# Express SAPS 3 per 10-point increase for easier interpretation.
adjusted_data$saps_3_10 <- adjusted_data$saps_3 / 10
```

# 6) Descriptive summary

Observed time is the minimum of event time and censoring time. It should not be interpreted as the distribution of potential follow-up. Median potential follow-up is estimated later with the reverse Kaplan-Meier method.

```{r descriptive-summary}
group_counts <- table(survival_data$group)
group_events <- tapply(
  survival_data$event,
  survival_data$group,
  sum
)

group_summary <- data.frame(
  Group = names(group_counts),
  Participants = as.integer(group_counts),
  Deaths = as.integer(group_events),
  Censored = as.integer(group_counts - group_events),
  Median_observed_time_days = as.numeric(tapply(
    survival_data$follow_up_days,
    survival_data$group,
    median
  ))
)

kable(
  group_summary,
  digits = 1,
  caption = "Descriptive summary by allocation group"
)
```
# 7) Kaplan-Meier curve, Kaplan-Meier curves by group, Numbers at risk and milestone survival, The absolute survival difference at time, and Median survival.

```{r survival-object}
survival_object <- with(
  survival_data,
  Surv(time = follow_up_days, event = event, type = "right")
)

head(survival_object)
```

## 7.1) Kaplan-Meier curve

```{r overall-kaplan-meier}
km_overall <- survfit(
  Surv(follow_up_days, event) ~ 1,
  data = survival_data,
  conf.type = "log-log"
)

summary(km_overall)

plot(
  km_overall,
  xlab = "Follow-up time (days)",
  ylab = "Estimated survival probability",
  main = "Overall Kaplan-Meier survival curve",
  mark.time = TRUE,
  conf.int = TRUE
)
```

## 7.2) Kaplan-Meier curves by group

```{r grouped-kaplan-meier}
km_by_group <- survfit(
  Surv(follow_up_days, event) ~ group,
  data = survival_data,
  conf.type = "log-log"
)

plot(
  km_by_group,
  xlab = "Follow-up time (days)",
  ylab = "Estimated survival probability",
  main = "Kaplan-Meier survival by allocation group",
  lty = 1:2,
  mark.time = TRUE,
  conf.int = FALSE
)

legend(
  "topright",
  legend = levels(survival_data$group),
  lty = 1:2,
  bty = "n"
)
```
## 7.3) Numbers at risk and milestone survival

```{r risk-table}
common_follow_up_limit <- min(tapply(
  survival_data$follow_up_days,
  survival_data$group,
  max
))

candidate_times <- c(0, 180, 365, 540)
risk_times <- candidate_times[candidate_times <= common_follow_up_limit]

risk_results <- summary(
  km_by_group,
  times = risk_times,
  extend = FALSE
)

risk_table <- data.frame(
  Group = sub("group=", "", risk_results$strata),
  Time_days = risk_results$time,
  Number_at_risk = risk_results$n.risk,
  Survival_probability = risk_results$surv,
  CI_lower = risk_results$lower,
  CI_upper = risk_results$upper
)

kable(
  risk_table,
  digits = 3,
  caption = "Numbers at risk and Kaplan-Meier survival estimates"
)
```
## 7.4) The absolute survival difference at time 

```{r milestone-survival}
milestone_day <- 365

if (milestone_day > common_follow_up_limit) {
  stop("The selected milestone exceeds the observed follow-up range of at least one group.")
}

milestone_results <- summary(
  km_by_group,
  times = milestone_day,
  extend = FALSE
)

milestone_table <- data.frame(
  Group = sub("group=", "", milestone_results$strata),
  Time_days = milestone_results$time,
  Survival_probability = milestone_results$surv,
  Standard_error = milestone_results$std.err,
  CI_lower = milestone_results$lower,
  CI_upper = milestone_results$upper
)

control_row <- match("Control", milestone_table$Group)
intervention_row <- match("Intervention", milestone_table$Group)

control_survival_365 <- milestone_table$Survival_probability[control_row]
intervention_survival_365 <- milestone_table$Survival_probability[intervention_row]

survival_difference_365 <-
  intervention_survival_365 - control_survival_365

survival_difference_se <- sqrt(
  milestone_table$Standard_error[intervention_row]^2 +
    milestone_table$Standard_error[control_row]^2
)

survival_difference_ci <- survival_difference_365 +
  c(-1, 1) * qnorm(0.975) * survival_difference_se

# A difference between two probabilities must lie between -1 and 1.
survival_difference_ci <- pmax(-1, pmin(1, survival_difference_ci))

kable(
  milestone_table,
  digits = 3,
  caption = "Estimated 365-day survival by allocation group"
)

survival_difference_365
survival_difference_ci
```

## 7.5) Median survival

Median survival is the earliest time at which the Kaplan-Meier estimate is 0.50 or lower.
It is not estimable if the curve does not reach 0.50 during observed follow-up.

```{r median-survival}
km_table <- summary(km_by_group)$table

median_survival_table <- data.frame(
  Group = sub("group=", "", rownames(km_table)),
  Median_days = km_table[, "median"],
  CI_lower = km_table[, "0.95LCL"],
  CI_upper = km_table[, "0.95UCL"]
)

kable(
  median_survival_table,
  digits = 1,
  caption = "Median survival by allocation group"
)
```

# 8) Log-rank test
The log-rank test compares observed and expected deaths over the full follow-up period.
The test provides a global unadjusted comparison but does not estimate an effect size.

```{r log-rank-test}
logrank_test <- survdiff(
  Surv(follow_up_days, event) ~ group,
  data = survival_data,
  rho = 0
)

logrank_df <- length(logrank_test$n) - 1
logrank_p_value <- pchisq(
  logrank_test$chisq,
  df = logrank_df,
  lower.tail = FALSE
)

logrank_test
logrank_p_value
```

# 9)  Unadjusted Cox proportional-hazards model
Under proportional hazards, an HR below 1 indicates a lower death hazard in the intervention group than in the control group among participants who remain alive at each time. It is not a fixed-time risk ratio or a ratio of survival probabilities.

```{r unadjusted-cox}
cox_unadjusted <- coxph(
  Surv(follow_up_days, event) ~ group,
  data = survival_data,
  ties = "efron",
  x = TRUE
)

unadjusted_summary <- summary(cox_unadjusted)
summary(cox_unadjusted)

unadjusted_hr <- unadjusted_summary$conf.int[
  "groupIntervention",
  "exp(coef)"
]

unadjusted_ci <- unadjusted_summary$conf.int[
  "groupIntervention",
  c("lower .95", "upper .95")
]

unadjusted_p_value <- unadjusted_summary$coefficients[
  "groupIntervention",
  "Pr(>|z|)"
]

# Unadjusted-cox-table
unadjusted_cox_table <- data.frame(
  Comparison = "Intervention vs Control",
  Hazard_ratio = unadjusted_hr,
  CI_lower = unadjusted_ci[1],
  CI_upper = unadjusted_ci[2],
  P_value = unadjusted_p_value
)

kable(
  unadjusted_cox_table,
  digits = 3,
  caption = "Unadjusted Cox proportional-hazards model"
)
```

## 9.1) Parsimonious adjusted Cox model
To preserve stability and clarity in this small educational dataset, the adjusted model includes allocation group and one prespecified baseline severity measure, SAPS 3. The SAPS 3 coefficient is scaled to a 10-point increase. The adjusted group coefficient is conditional on SAPS 3 and should not automatically be interpreted as causal.

```{r adjusted-cox}
cox_adjusted <- coxph(
  Surv(follow_up_days, event) ~ group + saps_3_10,
  data = adjusted_data,
  ties = "efron",
  x = TRUE
)

adjusted_summary <- summary(cox_adjusted)
summary(cox_adjusted)

# Adjusted-cox-table
adjusted_cox_table <- data.frame(
  Term = c(
    "Intervention vs Control",
    "SAPS 3, per 10-point increase"
  ),
  Hazard_ratio = adjusted_summary$conf.int[, "exp(coef)"],
  CI_lower = adjusted_summary$conf.int[, "lower .95"],
  CI_upper = adjusted_summary$conf.int[, "upper .95"],
  P_value = adjusted_summary$coefficients[, "Pr(>|z|)"]
)

kable(
  adjusted_cox_table,
  digits = 3,
  caption = "Parsimonious adjusted Cox proportional-hazards model"
)
```

## 9.2) Proportional-hazards assessment

Schoenfeld residual.
A systematic relationship between scaled Schoenfeld residuals and time suggests nonproportional hazards.
A nonsignificant test does not prove proportional hazards, particularly in a small sample. The covariate-specific and global tests should be interpreted with the residual plot and the Kaplan-Meier curves.

```{r proportional-hazards}
ph_test <- cox.zph(
  cox_adjusted,
  transform = "km"
)

ph_test

# The first model term is the allocation-group coefficient.
plot(
  ph_test,
  var = 1,
  xlab = "Follow-up time",
  ylab = "Scaled Schoenfeld residual",
  main = "Proportional-hazards assessment for allocation group"
)
abline(h = 0, lty = 2)

ph_group_p <- ph_test$table["group", "p"]
ph_global_p <- ph_test$table["GLOBAL", "p"]
```

# 10) Follow-up estimated by the reverse Kaplan-Meier method

The reverse Kaplan-Meier method reverses the death and censoring indicators.
Censoring times are treated as events and death times as censored observations. The resulting median is an estimate of median potential follow-up rather than the median of observed event-or-censoring times.

```{r reverse-kaplan-meier}
reverse_km <- survfit(
  Surv(follow_up_days, 1 - event) ~ 1,
  data = survival_data,
  conf.type = "log-log"
)

reverse_km_table <- summary(reverse_km)$table

median_follow_up_days <- unname(reverse_km_table["median"])
median_follow_up_ci <- unname(reverse_km_table[
  c("0.95LCL", "0.95UCL")
])

median_follow_up_days
median_follow_up_ci
```

# 11) Final summary table

```{r final-summary}
adjusted_group_hr <- adjusted_summary$conf.int[
  "groupIntervention",
  "exp(coef)"
]

adjusted_group_ci <- adjusted_summary$conf.int[
  "groupIntervention",
  c("lower .95", "upper .95")
]

adjusted_group_p <- adjusted_summary$coefficients[
  "groupIntervention",
  "Pr(>|z|)"
]

final_summary_table <- data.frame(
  Measure = c(
    "Log-rank test",
    "Unadjusted hazard ratio: intervention vs control",
    "SAPS 3-adjusted hazard ratio: intervention vs control",
    "365-day survival: control",
    "365-day survival: intervention",
    "Absolute 365-day survival difference",
    "Median potential follow-up"
  ),
  Estimate = c(
    sprintf("Chi-square = %.3f", logrank_test$chisq),
    sprintf("%.3f", unadjusted_hr),
    sprintf("%.3f", adjusted_group_hr),
    sprintf("%.1f%%", 100 * control_survival_365),
    sprintf("%.1f%%", 100 * intervention_survival_365),
    sprintf("%.1f %", 100 * survival_difference_365),
    sprintf("%.0f days", median_follow_up_days)
  ),
  `95% confidence interval` = c(
    "Not applicable",
    sprintf("%.3f to %.3f", unadjusted_ci[1], unadjusted_ci[2]),
    sprintf("%.3f to %.3f", adjusted_group_ci[1], adjusted_group_ci[2]),
    sprintf(
      "%.1f%% to %.1f%%",
      100 * milestone_table$CI_lower[control_row],
      100 * milestone_table$CI_upper[control_row]
    ),
    sprintf(
      "%.1f%% to %.1f%%",
      100 * milestone_table$CI_lower[intervention_row],
      100 * milestone_table$CI_upper[intervention_row]
    ),
    sprintf(
      "%.1f to %.1f percentage points",
      100 * survival_difference_ci[1],
      100 * survival_difference_ci[2]
    ),
    sprintf(
      "%.0f to %.0f days",
      median_follow_up_ci[1],
      median_follow_up_ci[2]
    )
  ),
  P_value = c(
    format.pval(logrank_p_value, digits = 3, eps = 0.001),
    format.pval(unadjusted_p_value, digits = 3, eps = 0.001),
    format.pval(adjusted_group_p, digits = 3, eps = 0.001),
    "Not applicable",
    "Not applicable",
    "Not calculated",
    "Not applicable"
  ),
  check.names = FALSE
)

kable(
  final_summary_table,
  caption = "Final summary of survival analyses"
)
```

## End of document
