---
title: "Reporting descriptive statistics in clinical research"
authors: "Renato Carneiro de Freitas Chaves, Tiago Mendonça dos Santos, Thiago Domingos Corrêa"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# 1) Purpose

This R Markdown document provides a step-by-step analytical workflow for clinical research comparing:
- **Control group** (`group = 0`)
- **Intervention group** (`group = 1`)

Using the attached dataset, it demonstrates how to:
1. Prepare an analysis-ready dataset for descriptive reporting.
2. Summarize **continuous variables** using measures of central tendency and dispersion.
3. Summarize **categorical variables** using counts and proportions.
4. Create tables and visual displays commonly used in epidemiological and clinical research.
5. Organize descriptive results in a format suitable for academic and scientific reporting.

The descriptive examples use:
- **Grouping variable**: `group`
- **Binary clinical outcome**: `outcome`
- **Continuous baseline/laboratory variables**, such as `age`, `platelet`, `hemoglobin`, `creatinine`, `saps_3`, and `fluid_balance`
- **Binary comorbidity/support variables**, such as `hypertension`, `diabetes`, `heart_failure`, `smoking`, `chronic_kidney_disease`, `cancer`, `stroke`, `vasopressor`, and `mechanical_ventilation`

> **Dataset coding (per the model data file)**  
> - `group`: `0 = Control`, `1 = Intervention`  
> - `outcome`: `0 = Alive`, `1 = Dead`

# 2) Packages

```{r packages}
# Install once if necessary:
# install.packages(c("readxl", "dplyr", "janitor", "tibble", "skimr", "ggplot2", "scales", "forcats", "gtsummary"))

library(readxl)
library(dplyr)
library(janitor)
library(tibble)
library(skimr)
library(ggplot2)
library(scales)
library(forcats)
library(gtsummary)
```

# 3) Data Import (Excel)

This guide uses the Excel file **data.xlsx** with a sheet named **"Data"**.

- If `data.xlsx` is in the same folder as this `.Rmd`, you can keep the path as `"data.xlsx"`.
- Otherwise, replace with the full path to your file.

```{r import-data}
data <- read_excel("data.xlsx", sheet = "Data")
```

# 4) Confirm Required Variables

For the requested descriptive analyses, we need:
- `group` (0 = Control, 1 = Intervention)
- `outcome` (0 = Alive, 1 = Dead)
- a set of continuous variables for numeric summaries
- a set of binary/categorical variables for frequency summaries

```{r verify-vars}
# Quick structure and missingness checks
skimr::skim(data)

# Example frequency check for the outcome variable
data %>% 
  tabyl(outcome)
```

# 5) Create the Analysis Dataset

We will:
- define the continuous and binary variables of interest
- keep only rows with non-missing `group` and `outcome`
- ensure continuous variables are numeric
- ensure binary variables are coded consistently
- create labeled factors for descriptive presentation

```{r analysis-dataset}
# --- Continuous variables ---
numerical_vars <- c(
  "age",
  "platelet",
  "hemoglobin",
  "creatinine",
  "saps_3",
  "fluid_balance"
)

# --- Binary/categorical variables ---
binary_vars <- c(
  "group",
  "hypertension",
  "diabetes",
  "heart_failure",
  "smoking",
  "chronic_kidney_disease",
  "cancer",
  "stroke",
  "vasopressor",
  "mechanical_ventilation",
  "outcome"
)

required_vars <- c(numerical_vars, binary_vars)

analysis_data <- data %>%
  select(all_of(required_vars)) %>%
  filter(!is.na(group), !is.na(outcome)) %>%
  mutate(
    across(all_of(numerical_vars), ~ as.numeric(.x)),
    across(all_of(binary_vars), ~ as.integer(.x))
  ) %>%
  mutate(
    group = factor(group, levels = c(0, 1), labels = c("Control", "Intervention")),
    outcome = factor(outcome, levels = c(0, 1), labels = c("Alive", "Dead")),
    across(
      all_of(setdiff(binary_vars, c("group", "outcome"))),
      ~ factor(.x, levels = c(0, 1), labels = c("No", "Yes"))
    )
  )

# Confirm final structure
str(analysis_data)
```

# 6) Descriptive Reporting for Categorical Variables

## 6.1 Frequency tables and proportions

This section summarizes the distribution of categorical variables using:
- **absolute frequencies (n)**
- **relative frequencies (%)**
- **contingency tables** when comparisons across groups are relevant

```{r categorical-distributions}
# Contingency table: Group vs Outcome
# Often central in clinical and epidemiological reports

tab_group_outcome <- table(analysis_data$group, analysis_data$outcome)
tab_group_outcome

# Row proportions: within-group outcome distribution
prop.table(tab_group_outcome, margin = 1)

# Example: Group vs hypertension

tab_group_htn <- table(analysis_data$group, analysis_data$hypertension)
tab_group_htn
prop.table(tab_group_htn, margin = 1)
```

# 7) Descriptive Reporting for Continuous Variables

This section summarizes continuous variables using:
- **mean** and **standard deviation (SD)**
- **median** and **interquartile range (IQR)**
- **minimum** and **maximum**

## 7.1 Continuous-variable summary by group

```{r continuous-summary}
continuous_summary <- analysis_data %>%
  group_by(group) %>%
  summarise(
    across(
      all_of(numerical_vars),
      list(
        n = ~ sum(!is.na(.x)),
        mean = ~ mean(.x, na.rm = TRUE),
        sd = ~ sd(.x, na.rm = TRUE),
        median = ~ median(.x, na.rm = TRUE),
        q1 = ~ quantile(.x, 0.25, na.rm = TRUE),
        q3 = ~ quantile(.x, 0.75, na.rm = TRUE),
        min = ~ min(.x, na.rm = TRUE),
        max = ~ max(.x, na.rm = TRUE)
      ),
      .names = "{.col}_{.fn}"
    ),
    .groups = "drop"
  )

continuous_summary
```

## 7.2 Stratified descriptive table for publication/reporting

```{r descriptive-table}
tab_01 <- analysis_data %>%
  select(all_of(c("group", setdiff(binary_vars, "group"), numerical_vars))) %>%
  tbl_summary(
    by = group,
    missing = "ifany",
    statistic = list(
      all_categorical() ~ "{n} ({p}%)",
      all_continuous()  ~ "{mean} ({sd}); {median} ({p25}, {p75}); min {min}, max {max}"
    ),
    digits = all_continuous() ~ 2
  ) %>%
  add_overall(last = TRUE) %>%
  bold_labels()

tab_01
```

# 8) Graphical Display of the Data

Visual summaries complement numerical summaries and help assess shape, skewness, variability, and outliers.

## 8.1 Histograms

```{r histograms, fig.height=4, fig.width=7}
for (v in numerical_vars) {

  p <- ggplot(analysis_data, aes(x = .data[[v]])) +
    geom_histogram(bins = 15, boundary = 0) +
    labs(
      title = paste("Histogram of", v),
      x = v,
      y = "Count"
    )

  print(p)
}
```

## 8.2 Density plots

```{r density-plots, fig.height=4, fig.width=7}
for (v in numerical_vars) {

  p <- ggplot(analysis_data, aes(x = .data[[v]])) +
    geom_density() +
    labs(
      title = paste("Density plot of", v),
      x = v,
      y = "Density"
    )

  print(p)
}
```

## 8.3 Box plots (overall and by group)

```{r box-plots, fig.height=4, fig.width=7}
for (v in numerical_vars) {

  # Overall box plot
  p1 <- ggplot(analysis_data, aes(y = .data[[v]])) +
    geom_boxplot() +
    labs(
      title = paste("Box plot of", v, "(overall)"),
      y = v,
      x = ""
    )

  print(p1)

  # By-group box plot
  p2 <- ggplot(analysis_data, aes(x = group, y = .data[[v]])) +
    geom_boxplot() +
    labs(
      title = paste("Box plot of", v, "by group"),
      x = "Group",
      y = v
    )

  print(p2)
}
```

# 9) Graphical Display for Categorical Variables

Pie charts can illustrate category composition, although bar charts are often preferred in scientific reporting.

## 9.1 Mathematical definition

For category \(c\), the slice size corresponds to the proportion \(p_c\), with percentage equal to \(100 \times p_c\).

```{r pie-charts, fig.height=4, fig.width=7}
# Reusable pie chart function for a factor variable
plot_pie <- function(df, var, title_prefix = "Pie chart of") {
  tab <- df %>%
    count(.data[[var]], name = "n") %>%
    mutate(
      proportion = n / sum(n),
      label = paste0(.data[[var]], "\n", percent(proportion, accuracy = 0.1))
    )

  ggplot(tab, aes(x = "", y = proportion, fill = .data[[var]])) +
    geom_col(width = 1) +
    coord_polar(theta = "y") +
    geom_text(aes(label = label), position = position_stack(vjust = 0.5), size = 3) +
    labs(
      title = paste(title_prefix, var),
      x = NULL,
      y = NULL,
      fill = var
    ) +
    theme_void()
}

# Example pie charts commonly reported
print(plot_pie(analysis_data, "group"))
print(plot_pie(analysis_data, "outcome"))
print(plot_pie(analysis_data, "hypertension"))
```

## End of document
