---
title: "Beyond statistical significance: interpretation and reporting of p values and confidence intervals with computational implementation in r and python"
authors: "Renato Carneiro de Freitas Chaves, Tiago Mendonça dos Santos, Thiago Domingos Corrêa"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# 1) Purpose

This R Markdown document provides a step-by-step analytical structure for clinical research comparing:
- **Control group** (`group = 0`)
- **Intervention group** (`group = 1`)

Using the attached dataset, it demonstrates how to:
1. Formulate **null and alternative hypotheses** for common epidemiological comparisons.
2. Choose and run **hypothesis tests** (with decision rules using **α**).
3. Compute and interpret **p values**.
4. Compute and interpret **95% confidence intervals (CIs)**, and connect CIs to hypothesis testing.

The example uses a **binary outcome** (`outcome`) compared between:
- **Control group** (`group = 0`)
- **Intervention group** (`group = 1`)

A secondary example illustrates hypothesis testing and CIs for a **continuous variable** (e.g., `age`) across groups.

> **Dataset coding (per the model data file)**  
> - `group`: `0 = Control`, `1 = Intervention`  
> - `outcome`: `0 = Alive`, `1 = Dead`  
> The **event of interest** is **death** (`outcome = 1`).

# 2) Packages

```{r packages}
# Install once if necessary:
# install.packages(c("readxl", "dplyr", "janitor", "tibble", "skimr"))

library(readxl)
library(dplyr)
library(janitor)
library(tibble)
library(skimr)
```

# 3) Data Import (Excel)
This guide uses the Excel file **data.xlsx** with a sheet named **"Data"**.

- If `data.xlsx` is in the same folder as this `.Rmd`, you can keep the path as `"data.xlsx"`.
- Otherwise, replace with the full path to your file.

```{r import-data}
data <- read_excel("data.xlsx", sheet = "Data") 

```


# 4) Confirm Required Variables

For the requested calculations, we need:
- `group` (0 = Control, 1 = Intervention)
- `outcome` (0 = Alive, 1 = Dead)

```{r verify-vars-2}
# Quick frequency checks
skimr::skim(data)

data %>% 
  tabyl(outcome)

## Creating a function that allows values smaller than 0.001 to be displayed as < 0.001
format_p <- function(p, digits = 3) {
  ifelse(
    is.na(p),
    NA,
    ifelse(
      p < 0.001,
      "< 0.001",
      formatC(p, format = "f", digits = digits)
    )
  )
}
```
# 5) Create the Analysis Dataset

We will:

- keep only rows with non-missing `group` **and** `outcome`
- ensure they are coded as expected (0/1)
- define the event as `outcome == 1` (death)

```{r analysis-dataset}
analysis_data <- data %>%
  select(group, outcome) %>%
  filter(!is.na(group), !is.na(outcome)) %>%
  mutate(
    group = as.integer(group),
    event = as.integer(outcome)  # event = 1 means "death"
  )

```
# 6) Construct the 2×2 Table

Rows:
- Intervention
- Control

Columns:
- Event (Dead)
- Non-event (Alive)

```{r table-2x2}
analysis_data <- analysis_data %>%
  mutate(
    group_label = factor(group, levels = c(1, 0), labels = c("Intervention", "Control")),
    event_label = factor(event, levels = c(1, 0), labels = c("Event", "Non-event"))
  )

tab <- table(analysis_data$group_label, analysis_data$event_label)
tab
```

# 7) Hypothesis Testing for a Binary Outcome 

## 7.1 Hypotheses, α level, and decision rule
### Null hypothesis (H0)
There is **no association** between `group` and `outcome` (event risk is equal across groups).

In probability notation:
\\[H_0: p_{Intervention} = p_{Control} \\]

### Alternative hypothesis (H1)
Event risk differs across groups:
\\[H_1: p_{Intervention} \\neq p_{Control}\\]

### Alpha (α)
We set a conventional two-sided significance level:
\\[\\alpha = 0.05\\]

**Decision rule**
- If **p value ≤ α**, reject *H0*.
- If **p value > α**, fail to reject *H0*.

```{r alpha}
alpha <- 0.05
alpha
```

## 7.2 Choosing the appropriate test (Chi-square vs Fisher's)
- **Chi-square test** is commonly used for 2×2 tables when expected counts are not too small.
- **Fisher’s exact test** is preferred when expected counts are small (a common rule of thumb: any expected cell count < 5).

```{r expected-counts}
# Expected counts for Chi-square suitability
chisq_expected <- suppressWarnings(chisq.test(tab, correct = FALSE)$expected)
chisq_expected
```

```{r run-tests}
# Run both tests (emphasize the recommended one)
## Chi-square test (no Yates continuity correction)
chisq_res <- suppressWarnings(chisq.test(tab, correct = FALSE))
p_value_chisq = chisq_res$p.value

## Fisher's exact test (exact p value)
fisher_res <- fisher.test(tab)
p_value_fisher = fisher_res$p.value

## Choose which p value to emphasize based on expected counts
use_fisher <- any(chisq_expected < 5)
test_recommended = ifelse(use_fisher, "Fisher's exact test", "Pearson Chi-square test")

```

## 7.3 Interpreting the p value in clinical/epidemiologic terms

```{r interpret-pvalue}
p_used <- if (use_fisher) fisher_res$p.value else chisq_res$p.value

decision <- if (p_used <= alpha) "Reject H0 " else "Fail to reject H0"

```
# 8) Hypothesis Testing for a Continuous Variable (Age)

This section demonstrates:
- Null/alternative hypotheses for means
- A two-sample t test (Welch) and its 95% CI

## 8.1 Hypotheses for age

\\[H_0: \\mu_{Intervention} = \\mu_{Control}\\]
\\[H_1: \\mu_{Intervention} \\neq \\mu_{Control}\\]

```{r continuous-prep}
# Descriptive summary of age

data <- data %>%
  mutate(group_label = factor(group,
                              levels = c(1,0),
                              labels = c("Intervention","Control")))

data %>% group_by(group_label) %>%
  summarise(
    n = n(),
    mean_age = mean(age, na.rm = TRUE),
    sd_age = sd(age, na.rm = TRUE),
    median_age = median(age, na.rm = TRUE),
    q1 = quantile(age, 0.25, na.rm = TRUE),
    q3 = quantile(age, 0.75, na.rm = TRUE),
    min = min(age, na.rm = TRUE),
    max = max(age, na.rm = TRUE)
  )


```

## 8.2 Welch two-sample t test (default in `t.test`)

`Welch` is preferred when variances may differ.

```{r ttest-age}
tt_age <- t.test(age ~ group_label, data = data, conf.level = 1 - alpha)

p_value = tt_age$p.value
mean_difference = unname(diff(rev(tt_age$estimate))) # Intervention - Control (depends on factor order)
CI_low = tt_age$conf.int[1]
CI_high = tt_age$conf.int[2]
  
```

# 9) Final Summary Table (Binary Outcome)

```{r final-summary}
final_summary_table <- tibble::tribble(~measure, ~value,                                   ~lower_ci, ~upper_ci,
  "Recommended test for group vs outcome",       test_recommended,                          NA,        NA,
  "P value (Chi-square); group vs outcome",      format_p(p_value_chisq),                   NA,        NA,
  "P value (Fisher); group vs outcome",          format_p(p_value_fisher),                  NA,        NA,
  "Alpha (α); group vs outcome",                 formatC(alpha, format="f", digits=2),      NA,        NA,
  "Decision at α=0.05; group vs outcome",        decision,                                  NA,        NA,
  "P value for age (group comparison)",          format_p(p_value),                          NA,        NA,
  "Age mean difference (group comparison)",      formatC(mean_difference, format="f", digits=2),
                                                 formatC(CI_low, format="f", digits=2),
                                                 formatC(CI_high, format="f", digits=2)
)
    
final_summary_table
```

---

## End of document
