---
title: "Risk, Relative Risk, Odds, and Odds Ratio"
authors: "Renato Carneiro de Freitas Chaves, Tiago Mendonça dos Santos, Thiago Domingos Corrêa"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# Purpose

This R Markdown document provides a step-by-step analytical guide for calculating and interpreting four fundamental epidemiological measures:

1. **Risk**
2. **Relative risk**
3. **Odds**
4. **Odds ratio**

The analysis uses a binary exposure variable and a binary outcome variable from the attached dataset.

> **Dataset coding**  
> - `group`: `0 = Control`, `1 = Intervention`  
> - `outcome`: `0 = Alive`, `1 = Dead`  
>
> The event of interest in this report is **death** (`outcome = 1`).

# 2) Packages

```{r packages}
# Install these packages once if necessary:
# install.packages(c("readxl", "dplyr", "janitor", "tibble", "knitr", "skimr"))

library(readxl)
library(dplyr)
library(janitor)
library(tibble)
library(knitr)
library(skimr)
```

# 3) Data Import (Excel)
This guide uses the Excel file **data.xlsx** with a sheet named **"Data"**.

- If `data.xlsx` is in the same folder as this `.Rmd`, you can keep the path as `"data.xlsx"`.
- Otherwise, replace with the full path to your file.

```{r data-loading}

data <- read_excel("data.xlsx", sheet = "Data") 

```

# 4) Confirm Required Variables

For the requested calculations, we need:
- `group` (0 = Control, 1 = Intervention)
- `outcome` (0 = Alive, 1 = Dead)

```{r data-inspection}
# Quick frequency checks
skimr::skim(data)

data %>% 
  tabyl(outcome)

```
# 5) Create the Analysis Dataset
We will:
- keep only rows with non-missing `group` **and** `outcome`
- ensure they are coded as expected (0/1)
- define the event as `outcome == 1` (death)


- **Exposure variable**: `group`
  - `group = 1`: Intervention group
  - `group = 0`: Control group

- **Outcome variable**: `outcome`
  - `outcome = 1`: Death
  - `outcome = 0`: Alive
  
```{r data-preparation}
# Create an analysis dataset containing only the variables needed for the calculations.

analysis_data <- data %>%
  select(group, outcome) %>%
  filter(!is.na(group), !is.na(outcome)) %>%
  mutate(
    group = as.integer(group),
    outcome = as.integer(outcome),
    event = outcome,
    group_label = factor(
      group,
      levels = c(1, 0),
      labels = c("Intervention", "Control")
    ),
    event_label = factor(
      event,
      levels = c(1, 0),
      labels = c("Death", "Alive")
    )
  )

```

# 6) Construct the 2×2 Table

Rows:
- Intervention
- Control

Columns:
- Event (Dead)
- Non-event (Alive)

```{r table-2x2}
tab <- table(
  analysis_data$group_label,
  analysis_data$event_label
)
tab


# Extract the four cell counts using epidemiological notation.
a <- as.numeric(tab["Intervention", "Death"])  # Deaths in the intervention group
b <- as.numeric(tab["Intervention", "Alive"])  # Survivors in the intervention group
c <- as.numeric(tab["Control", "Death"])       # Deaths in the control group
d <- as.numeric(tab["Control", "Alive"])       # Survivors in the control group

# Define group totals.
n_intervention <- a + b      # Total number of patients in the intervention group
n_control <- c + d    # Total number of patients in the control group
n_total <- n_intervention + n_control

```

# 7) Risk

Risk represents the probability of hospital death during the observed hospitalization period.

For the intervention group:
\[Risk_{intervention} = \frac{a}{a+b}\]

For the control group:
\[Risk_{control} = \frac{c}{c+d}\]

```{r risk-calculation}
# Calculate the risk of death in the intervention group.
risk_intervention <- a / (a + b)

# Calculate the risk of death in the control group.
risk_control <- c / (c + d)

# Convert risks to percentages for clinical interpretation.
risk_intervention_percent <- risk_intervention * 100
risk_control_percent <- risk_control * 100
```
# 8) Relative Risk

Relative risk, also called the risk ratio, compares the risk of the event in the intervention group with the risk of the event in the control group.

\[Relative\ Risk = RR = \frac{Risk_{intervention}}{Risk_{control}}\]

Using the 2 × 2 table notation:

\[RR = \frac{a/(a+b)}{c/(c+d)}\]

```{r relative-risk-calculation}
# Calculate the relative risk.
relative_risk <- risk_intervention / risk_control

```

# 9) Odds
Odds express the probability of the event divided by the probability of the event not occurring.

For the intervention group:

\[Odds_{intervention} = \frac{a}{b}\]

For the control group:

\[Odds_{control} = \frac{c}{d}\]

```{r odds-calculation}
# Calculate the odds of death in the intervention group.
odds_intervention <- a / b

# Calculate the odds of death in the control group.
odds_control <- c / d

```

# 10) Odds Ratio
The odds ratio compares the odds of the event in the intervention group with the odds of the event in the control group.
\[Odds\ Ratio = OR = \frac{Odds_{intervention}}{Odds_{control}}\]

Using the 2 × 2 table notation:
\[OR = \frac{a/b}{c/d}\]

This can also be written as:

\[OR = \frac{a \times d}{b \times c}\]

```{r odds-ratio-calculation}
# Calculate the odds ratio directly from the odds.
odds_ratio <- odds_intervention / odds_control

# Calculate the same odds ratio using the cross-product formula.
odds_ratio_cross_product <- (a * d) / (b * c)

```

# Final Summary

```{r final-summary}
final_summary_table <- tibble(
  Measure = c(
    "Risk of death in intervention group",
    "Risk of death in control group",
    "Relative risk",
    "Odds of death in intervention group",
    "Odds of death in control group",
    "Odds ratio"
  ),
  Formula = c(
    "a / (a + b)",
    "c / (c + d)",
    "[a / (a + b)] / [c / (c + d)]",
    "a / b",
    "c / d",
    "(a / b) / (c / d) = (a × d) / (b × c)"
  ),
  Estimation = c(
    risk_intervention,
    risk_control,
    relative_risk,
    odds_intervention,
    odds_control,
    odds_ratio
  ),
  Interpretation = c(
    "Probability of death among patients in the intervention group.",
    "Probability of death among patients in the control group.",
    "Probability of death among patients in the intervention group compares to the probability in the control group.",
    "Odds of death relative to survival in the intervention group.",
    "Odds of death relative to survival in the control group.",
    "Odds of death odds in the intervention group compares to the odds in the control group."
  )
)

final_summary_table %>%
  kable(
    digits = 2,
    caption = "Reporting risk and odds in clinical research"
  )
```

---

## End of document
