
title: "Correlation and linear regression in clinical and public health research with computational implementation in r and python"
authors: "Renato Carneiro de Freitas Chaves, Tiago Mendonça dos Santos, Thiago Domingos Corrêa"
date: "`r Sys.Date()`"
output: html_document


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# 1) Purpose

This R Markdown document provides a clear, step-by-step analytical guide for evaluating the relationship between two continuous clinical variables using **correlation** and **linear regression**.

The applied objective is to use the attached clinical dataset to answer the following questions:

1. Is patient age linearly associated with serum creatinine at ICU admission?
2. How much does mean serum creatinine change, on average, for each additional year of age?
3. How should the statistical estimates be interpreted in clinical and epidemiological terms?


# 2) Packages

```{r packages}
# Install once if necessary:
# install.packages("readxl")

library(readxl)
```

# 3) Data Import (Excel)
This guide uses the Excel file **data.xlsx** with a sheet named **"Data"**.

- If `data.xlsx` is in the same folder as this `.Rmd`, you can keep the path as `"data.xlsx"`.
- Otherwise, replace with the full path to your file.

```{r import-data}
data <- read_excel("data.xlsx", sheet = "Data") 

```
# 4)  Variable Definitions and Data Inspection
Before conducting statistical analysis, the investigator should inspect the structure of the dataset, confirm the variable names, and review the first rows.


The main variables used in this report are:
- `age`: patient age at ICU admission, measured in years.
- `creatinine`: serum creatinine level, measured in mg/dL.
- `saps_3`: Simplified Acute Physiology Score 3 at ICU admission.
- `chronic_kidney_disease`: indicator of chronic kidney dysfunction, coded as 0 = no and 1 = yes.
- `vasopressor`: use of any vasopressor at ICU admission, coded as 0 = no and 1 = yes.
- `outcome`: vital status at hospital discharge, coded as 0 = alive and 1 = dead.

```{r inspect-data}
# Number of rows and columns.
dim(data)

# Variable names.
names(data)

# First six records.
head(data)
```

# 5) Basic Data Preparation

The primary analysis requires complete observations for `age` and `creatinine`. These variables are converted to numeric format to avoid problems caused by accidental text import.

```{r prepare-data}
# Start from the imported dataset.
analysis_data <- data

# Convert the continuous variables to numeric values.
analysis_data$age <- as.numeric(analysis_data$age)
analysis_data$creatinine <- as.numeric(analysis_data$creatinine)
analysis_data$saps_3 <- as.numeric(analysis_data$saps_3)

# Convert binary clinical indicators to numeric values first.
# This step is useful if the variables were imported as text.
analysis_data$chronic_kidney_disease <- as.numeric(analysis_data$chronic_kidney_disease)
analysis_data$vasopressor <- as.numeric(analysis_data$vasopressor)

# Remove records with missing values in any variable used in the analyses.
# Applying complete-case selection after numeric conversion prevents accidental
# retention of values that become missing during conversion.
analysis_data <- analysis_data[
  complete.cases(analysis_data[, c(
    "age",
    "creatinine",
    "saps_3",
    "chronic_kidney_disease",
    "vasopressor"
  )]),
]

# Convert binary indicators to factors for clearer regression output.
# The reference category is "No" for both variables.
analysis_data$chronic_kidney_disease <- factor(
  analysis_data$chronic_kidney_disease,
  levels = c(0, 1),
  labels = c("No", "Yes")
)

analysis_data$vasopressor <- factor(
  analysis_data$vasopressor,
  levels = c(0, 1),
  labels = c("No", "Yes")
)

# Confirm the analytic sample size and the observed age range.
n <- nrow(analysis_data)
n
range(analysis_data$age)

# Summarize the two primary continuous variables.
summary(analysis_data[, c("age", "creatinine")])
```

# 6) Graphical inspection before modeling

A scatterplot should precede correlation and regression.

```{r scatterplot, fig.width=7, fig.height=5}
plot(
  analysis_data$age,
  analysis_data$creatinine,
  xlab = "Age at ICU admission (years)",
  ylab = "Serum creatinine (mg/dL)",
  main = "Scatterplot of age and serum creatinine",
  pch = 19
)
```

# 7) Pearson Correlation
The value of the Pearson correlation coefficient ranges from -1 to +1.
Values near +1 indicate a strong positive linear association, values near -1 indicate a strong negative linear association, and values near 0 indicate little or no linear association.
Calculation of Correlation Test and Confidence Interval are described above


```{r correlation-test}
correlation_result <- cor.test(
  analysis_data$age,
  analysis_data$creatinine,
  method = "pearson",
  conf.level = 0.95
)

correlation_result
```

## 8 Interpretation of correlation
Clinical interpretation should focus on direction, magnitude, uncertainty, and plausibility. A positive value suggests that older patients tend to have higher creatinine values, whereas a negative value suggests that older patients tend to have lower creatinine values. A correlation close to zero does not exclude nonlinear relationships or subgroup-specific patterns.

```{r correlation-summary}
r_value <- unname(correlation_result$estimate)
r_ci_low <- correlation_result$conf.int[1]
r_ci_high <- correlation_result$conf.int[2]
r_p_value <- correlation_result$p.value

correlation_summary <- data.frame(
  measure = c("Pearson correlation coefficient", "95% CI lower bound", "95% CI upper bound", "P value"),
  value = c(r_value, r_ci_low, r_ci_high, r_p_value)
)

knitr::kable(correlation_summary, digits = 3, caption = "Pearson correlation between age and creatinine")
```
# 9) Fitting the Regression Model with `lm()`

Simple linear regression models the conditional mean of a continuous outcome:
\[E(Y_i|X_i) = \beta_0 + \beta_1X_i\]

For the present analysis:
\[E(\text{creatinine}_i|\text{age}_i) = \beta_0 + \beta_1 \times \text{age}_i\]

```{r simple-regression}
simple_model <- lm(creatinine ~ age, data = analysis_data)

summary(simple_model)
```

## 10 Regression Coefficients and 95% Confidence Intervals


```{r simple-regression-ci}
# Coefficient table from the fitted model.
coef_table <- summary(simple_model)$coefficients

# 95% confidence intervals for the intercept and slope.
coef_ci <- confint(simple_model, level = 0.95)

# Combine estimates and confidence intervals in one table.
simple_regression_results <- data.frame(
  term = rownames(coef_table),
  estimate = coef_table[, "Estimate"],
  standard_error = coef_table[, "Std. Error"],
  p_value = coef_table[, "Pr(>|t|)"],
  ci_low = coef_ci[, 1],
  ci_high = coef_ci[, 2]
)

knitr::kable(simple_regression_results, digits = 4, caption = "Simple linear regression: creatinine as a function of age")
```

## 11 Coefficient of determination

The coefficient of determination is:

\[R^2 = 1 - \frac{\sum_{i=1}^{n}(Y_i - \hat{Y}_i)^2}{\sum_{i=1}^{n}(Y_i - \bar{Y})^2}\]

where:

- \(\hat{Y}_i\) is the fitted creatinine value for patient \(i\).
- \(R^2\) is the proportion of observed variability in creatinine explained by the model.

In simple linear regression with one predictor and an intercept, \(R^2 = r^2\).

```{r r-squared}
r_squared_model <- summary(simple_model)$r.squared
r_squared_from_correlation <- r_value^2

r_squared_summary <- data.frame(
  measure = c("R-squared from regression", "Squared Pearson correlation"),
  value = c(r_squared_model, r_squared_from_correlation)
)

knitr::kable(r_squared_summary, digits = 4, caption = "Relationship between R-squared and Pearson correlation")
```

# 12) Fitted values and residuals

```{r fitted-residuals}
analysis_data$fitted_creatinine <- fitted(simple_model)
analysis_data$residual <- resid(simple_model)

head(analysis_data[, c("patient", "age", "creatinine", "fitted_creatinine", "residual")])
```

```{r regression-line, fig.width=7, fig.height=5}
plot(
  analysis_data$age,
  analysis_data$creatinine,
  xlab = "Age at ICU admission (years)",
  ylab = "Serum creatinine (mg/dL)",
  main = "Linear regression of creatinine on age",
  pch = 19
)
abline(simple_model, lwd = 2)
```

# 13) Model Diagnostics
Linear regression should not be interpreted only from the coefficient table.
The following diagnostic plots evaluate model assumptions and influential observations.

Interpretation of diagnostic plots:
- The residuals-versus-fitted plot helps assess nonlinearity and unequal variance.
- The normal Q-Q plot helps evaluate whether residuals are approximately normally distributed.
- The scale-location plot helps evaluate heteroscedasticity.
- The residuals-versus-leverage plot helps identify influential observations.


```{r diagnostics, fig.width=7, fig.height=7}
par(mfrow = c(2, 2))
plot(simple_model)
par(mfrow = c(1, 1))
```
# 14) Adjusted Linear Regression

Clinical and public health analyses often require adjustment for other measured factors.
In this example, creatinine is modeled as a function of age, severity of illness, chronic kidney disease, and vasopressor use.
The binary variables `chronic_kidney_disease` and `vasopressor` are represented as factors with `No` as the reference category.

```{r adjusted-regression}
adjusted_model <- lm(
  creatinine ~ age + saps_3 + chronic_kidney_disease + vasopressor,
  data = analysis_data
)

summary(adjusted_model)
```

```{r adjusted-regression-table}
adj_coef_table <- summary(adjusted_model)$coefficients
adj_coef_ci <- confint(adjusted_model, level = 0.95)

adjusted_regression_results <- data.frame(
  term = rownames(adj_coef_table),
  estimate = adj_coef_table[, "Estimate"],
  standard_error = adj_coef_table[, "Std. Error"],
  p_value = adj_coef_table[, "Pr(>|t|)"],
  ci_low = adj_coef_ci[, 1],
  ci_high = adj_coef_ci[, 2]
)

knitr::kable(adjusted_regression_results, digits = 4, caption = "Adjusted linear regression results")
```

# 15) Predicted Mean Creatinine for Selected Ages

These predictions should be interpreted only within the observed or clinically plausible range of age in the dataset.
The code below uses confidence intervals for the expected mean creatinine, not prediction intervals for individual patients.

These intervals estimate uncertainty around the expected mean creatinine at each selected age.
They should not be interpreted as the expected range for an individual patient's creatinine value. For individual-level prediction, the argument `interval = "prediction"` would be required.


```{r predictions}
new_patients <- data.frame(age = c(40, 60, 80))

predicted_creatinine <- predict(
  simple_model,
  newdata = new_patients,
  interval = "confidence",
  level = 0.95
)

prediction_table <- cbind(new_patients, predicted_creatinine)
knitr::kable(prediction_table, digits = 3, caption = "Predicted mean creatinine at selected ages")

# Example only: prediction intervals for individual patients would be requested as follows.
predict(simple_model, newdata = new_patients, interval = "prediction", level = 0.95)
```
# 16) Final Summary Table

```{r final-summary}
format_p <- function(p) {
  ifelse(p < 0.001, "<0.001", sprintf("%.3f", p))
}

slope_age <- simple_regression_results[simple_regression_results$term == "age", ]
adj_slope_age <- adjusted_regression_results[adjusted_regression_results$term == "age", ]

final_summary_table <- data.frame(
  Analysis = c(
    "Pearson correlation: age and creatinine",
    "Simple regression slope: creatinine on age",
    "Simple regression R-squared",
    "Adjusted regression slope for age"
  ),
  Estimate = c(
    sprintf("%.3f", r_value),
    sprintf("%.4f", slope_age$estimate),
    sprintf("%.3f", r_squared_model),
    sprintf("%.4f", adj_slope_age$estimate)
  ),
  Confidence_interval = c(
    paste0("95% CI ", sprintf("%.3f", r_ci_low), " to ", sprintf("%.3f", r_ci_high)),
    paste0("95% CI ", sprintf("%.4f", slope_age$ci_low), " to ", sprintf("%.4f", slope_age$ci_high)),
    "Not applicable",
    paste0("95% CI ", sprintf("%.4f", adj_slope_age$ci_low), " to ", sprintf("%.4f", adj_slope_age$ci_high))
  ),
  P_value = c(
    format_p(r_p_value),
    format_p(slope_age$p_value),
    "Not applicable",
    format_p(adj_slope_age$p_value)
  )
)

knitr::kable(final_summary_table, caption = "Final summary of correlation and linear regression analyses")
```

## End of document
