
title: "Diagnostic Accuracy Measures in Clinical Epidemiology"
authors: "Renato Carneiro de Freitas Chaves, Tiago Mendonça dos Santos, Thiago Domingos Corrêa"
date: "`r Sys.Date()`"
output: html_document

```{r setup}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# 1) Purpose

This R Markdown document provides a step-by-step analytical guide for estimating and interpreting key measures of diagnostic and prognostic performance in clinical epidemiology.

Using the attached dataset, this report demonstrates how to calculate and interpret:

1. **Sensitivity**
2. **Specificity**
3. **Positive and negative predictive values**
4. **Positive and negative likelihood ratios**
5. **Area under the receiver operating characteristic curve (AUC)**

This report estimates diagnostic/prognostic performance measures using:

- `outcome`: reference outcome (`0 = alive`, `1 = death`);
- `vasopressor`: binary index marker (`0 = no`, `1 = yes`);
- `saps_3`: continuous severity score for ROC/AUC analysis.

The event of interest is **hospital death**.
Because vasopressor use is not a diagnostic test in the usual sense, the binary measures should be interpreted as **prognostic classification measures** for hospital death, not as evidence that vasopressors cause death.

# 2) Packages

```{r packages}
# Install these packages once if necessary:
# install.packages(c("readxl", "dplyr", "janitor", "tibble", "knitr", "pROC", "skimr"))

library(readxl)
library(dplyr)
library(janitor)
library(tibble)
library(knitr)
library(pROC)
library(skimr)
```

# 3) Data Import (Excel)
This guide uses the Excel file **data.xlsx** with a sheet named **"Data"**.

- If `data.xlsx` is in the same folder as this `.Rmd`, you can keep the path as `"data.xlsx"`.
- Otherwise, replace with the full path to your file.

```{r import-data}

data <- read_excel("data.xlsx", sheet = "Data") 

```

# 4) Confirm Required Variables

For the requested calculations, we need:
- `outcome` (0 = Alive, 1 = Dead)
- `vasopressor` (0 = no, 1 = yes)


```{r verify-vars-2}
# Quick frequency checks
skimr::skim(data)

data %>% 
  tabyl(outcome)

data %>% 
  tabyl(vasopressor)

```

# 5) Create the Analysis Dataset

Define the variables used in this analysis
- `reference_var`: the true disease or outcome status. Here, hospital death is the event of interest.
- `binary_test_var`: the binary index test or clinical indicator evaluated against the reference outcome.
- `continuous_marker_var`: the continuous marker used to estimate ROC/AUC.

```{r define-variables}
# Reference standard:
disease <- data$outcome
test <- data$vasopressor
score <- as.numeric(data$saps_3)


# Create the analytical dataset
analysis_data <- data.frame(disease, test)
analysis_data <- analysis_data[complete.cases(analysis_data), ]

```

# 6) Construct the 2×2 Table

For a binary test evaluated against a binary reference standard, the data can be represented as follows:
- **TP** = true positives: test positive and disease positive
- **FP** = false positives: test positive and disease negative
- **FN** = false negatives: test negative and disease positive
- **TN** = true negatives: test negative and disease negative

```{r contingency-table}
# Construct the 2 x 2 table.
tab <- table(
  Test = factor(analysis_data$test, levels = c(1, 0), labels = c("Positive", "Negative")),
  Disease = factor(analysis_data$disease, levels = c(1, 0), labels = c("Present", "Absent"))
)
tab
# Extract the four fundamental cells of the 2 x 2 table.
TP <- tab["Positive", "Present"]
FP <- tab["Positive", "Absent"]
FN <- tab["Negative", "Present"]
TN <- tab["Negative", "Absent"]

```

# 7) Step-by-Step Calculation of Diagnostic Accuracy Measures

## 7.1 Sensitivity

Sensitivity is the probability that the test is positive among individuals who truly have the condition or outcome.

\[\text{Sensitivity} = \frac{TP}{TP + FN}\]

```{r sensitivity}
sensitivity <- TP / (TP + FN)
sensitivity
```

## 7.2 Specificity

Specificity is the probability that the test is negative among individuals who truly do not have the condition or outcome.

\[\text{Specificity} = \frac{TN}{TN + FP}\]

```{r specificity}
specificity <- TN / (TN + FP)
specificity
```

## 7.3 Positive Predictive Value

The positive predictive value is the probability of having the condition or outcome among individuals with a positive test.

\[\text{PPV} = \frac{TP}{TP + FP}\]

```{r ppv}
positive_predictive_value <- TP / (TP + FP)
positive_predictive_value
```

## 7.4 Negative Predictive Value

The negative predictive value is the probability of not having the condition or outcome among individuals with a negative test.

\[\text{NPV} = \frac{TN}{TN + FN}\]

```{r npv}
negative_predictive_value <- TN / (TN + FN)
negative_predictive_value
```

## 7.5 Positive Likelihood Ratio

The positive likelihood ratio indicates how many times more likely a positive test result is among individuals with the condition than among individuals without the condition.

\[\text{LR+} = \frac{\text{Sensitivity}}{1 - \text{Specificity}}\]

```{r lr-positive}
positive_likelihood_ratio <- sensitivity / (1 - specificity)
positive_likelihood_ratio
```

## 7.6 Negative Likelihood Ratio

The negative likelihood ratio indicates how many times less likely a negative test result is among individuals with the condition than among individuals without the condition.

\[\text{LR-} = \frac{1 - \text{Sensitivity}}{\text{Specificity}}\]

```{r lr-negative}
negative_likelihood_ratio <- (1 - sensitivity) / specificity
negative_likelihood_ratio
```

# 8) Summary Table of Diagnostic Accuracy Measures

```{r final-summary}

format_percent <- function(x, digits = 1) {
  ifelse(is.na(x), NA_character_, paste0(round(100 * x, digits), "%"))
}

format_number <- function(x, digits = 2) {
  ifelse(is.na(x), NA_character_, format(round(x, digits), nsmall = digits))
}

final_summary_table <- tibble(
  Measure = c(
    "Sensitivity",
    "Specificity",
    "Positive predictive value",
    "Negative predictive value",
    "Positive likelihood ratio",
    "Negative likelihood ratio"
  ),
  Formula = c(
    "TP / (TP + FN)",
    "TN / (TN + FP)",
    "TP / (TP + FP)",
    "TN / (TN + FN)",
    "Sensitivity / (1 - Specificity)",
    "(1 - Sensitivity) / Specificity"
  ),
  Estimate = c(
    format_percent(sensitivity),
    format_percent(specificity),
    format_percent(positive_predictive_value),
    format_percent(negative_predictive_value),
    format_number(positive_likelihood_ratio),
    format_number(negative_likelihood_ratio)
  )
)

kable(final_summary_table, digits = 3)
```

# 9) ROC Curve and Area Under the Curve

```{r auc}
# Keep only complete observations for the outcome and continuous predictor. Disease: binary outcome; 0 = non-event, 1 = event.
roc_data <- data.frame(
  disease = disease,
  score = score)
roc_data <- roc_data[complete.cases(roc_data), ]

# Estimate the ROC curve.
roc_obj <- pROC::roc(
  response = roc_data$disease,
  predictor = roc_data$score,
  levels = c(0, 1),
  direction = "<",
  ci = TRUE,
  quiet = TRUE
)

# Extract AUC and 95% confidence interval.
AUC = round(as.numeric(pROC::auc(roc_obj)), 3)
CI_95_lower = round(as.numeric(roc_obj$ci[1]), 3)
CI_95_upper = round(as.numeric(roc_obj$ci[3]), 3)




# Plot ROC curve.
plot(
  roc_obj,
  ylim = c(0, 1),
  asp = 1,
  col = "#2C7BB6",
  lwd = 2,
  grid = TRUE,
  main = "ROC Curve: SAPS 3 Predicting Hospital Death",
  legacy.axes = TRUE,
  print.auc = TRUE
)


abline(a = 0, b = 1, lty = 2, col = "gray70")
```

## End of document