---
title: "Calculating adequate sample size across clinical scenarios with computational implementation in r and python"
authors: "Renato Carneiro de Freitas Chaves, Tiago Mendonça dos Santos, Thiago Domingos Corrêa"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# 1) Purpose

This document presents a practical R Markdown document for sample-size calculations in common clinical and epidemiological scenarios.

The examples use a small number of parameters that can be edited directly by the researcher.


# 2) Packages
```{r packages}
# Install once if necessary:
# install.packages(c("readxl", "dplyr", "tibble", "janitor", "knitr"))

library(readxl)
library(dplyr)
library(tibble)
library(janitor)
library(knitr)
```

# 3) Data Import (Excel)
This guide uses the Excel file **data.xlsx** with a sheet named **"Data"**.

- If `data.xlsx` is in the same folder as this `.Rmd`, you can keep the path as `"data.xlsx"`.
- Otherwise, replace with the full path to your file.


- `group`: 0 = control, 1 = intervention
- `outcome`: 0 = no event, 1 = event

```{r import-data}

data <- read_excel("data.xlsx", sheet = "Data") 

```

# 4) User-defined parameters

Edit this block according to your study.

```{r parameters}
# General parameters
alpha <- 0.05
power <- 0.80
attrition <- 0.10


group_var <- "group"       # 0 = Control, 1 = Intervention
outcome_var <- "outcome"   # 0 = No event, 1 = Event
continuous_var <- "continuous_outcome"

# Example assumptions used below
precision_prevalence <- 0.05
precision_mean <- 3
minimal_difference_mean <- 5

p_control_default <- 0.25
p_intervention_default <- 0.15
sd_default <- 10

odds_ratio <- 2.0
p_exposed_controls <- 0.20

sensitivity <- 0.85
specificity <- 0.90
precision_sensitivity <- 0.07
precision_specificity <- 0.05
disease_prevalence <- 0.20

```


# 5) Simple descriptive estimates from the dataset

These estimates are optional. They are useful for demonstration, but the final assumptions should be clinically justified before being used in a study protocol.

```{r descriptive-estimates}
overall_event_risk <- mean(data$outcome, na.rm = TRUE)

risk_by_group <- data %>%
  group_by(group) %>%
  summarise(
    n = n(),
    event_risk = mean(outcome, na.rm = TRUE),
    .groups = "drop"
  )

kable(risk_by_group, digits = 3, caption = "Observed event risk by group")
```

```{r assumptions-from-data}
# Data-informed values for the examples.
# Replace these with protocol-defined assumptions whenever appropriate.

risk_control <- risk_by_group$event_risk[match(0, risk_by_group$group)]
risk_intervention <- risk_by_group$event_risk[match(1, risk_by_group$group)]

risk_control <- coalesce(risk_control, p_control_default)
risk_intervention <- coalesce(risk_intervention, p_intervention_default)

continuous_sd <- sd(data[[continuous_var]], na.rm = TRUE)
continuous_sd <- coalesce(continuous_sd, sd_default)

assumptions <- tibble(
  parameter = c(
    "overall_event_risk",
    "risk_control",
    "risk_intervention",
    "continuous_sd",
    "alpha",
    "power",
    "attrition"
  ),
  value = c(
    overall_event_risk,
    risk_control,
    risk_intervention,
    continuous_sd,
    alpha,
    power,
    attrition
  )
)

kable(assumptions, digits = 3, caption = "Main assumptions used in the examples")
```

# 6) Useful constants

```{r constants}
z_alpha <- qnorm(1 - alpha / 2)
z_power <- qnorm(power)

constants <- tibble(
  parameter = c("Z for alpha", "Z for power"),
  value = c(z_alpha, z_power)
)

kable(constants, digits = 3, caption = "Statistical constants")
```

# 7) Scenario 1: prevalence survey

For a single proportion estimated with absolute precision:

\[n = \frac{Z_{1-\alpha/2}^{2}p(1-p)}{d^2}\]

```{r prevalence-survey}
p <- overall_event_risk
d <- precision_prevalence

n_prevalence <- z_alpha^2 * p * (1 - p) / d^2
n_prevalence_final <- ceiling(n_prevalence / (1 - attrition))

prevalence_result <- tibble(
  scenario = "Prevalence survey",
  expected_prevalence = p,
  precision = d,
  sample_size_before_attrition = ceiling(n_prevalence),
  final_sample_size = n_prevalence_final
)

kable(prevalence_result, digits = 3, caption = "Sample size for estimating a prevalence")
```

# 8) Scenario 2: estimating a single mean

For a single mean estimated with absolute precision:

\[n = \frac{Z_{1-\alpha/2}^{2}\sigma^2}{d^2}\]

```{r single-mean}
sigma <- continuous_sd
d <- precision_mean

n_single_mean <- z_alpha^2 * sigma^2 / d^2
n_single_mean_final <- ceiling(n_single_mean / (1 - attrition))

single_mean_result <- tibble(
  scenario = "Single mean",
  expected_sd = sigma,
  precision = d,
  sample_size_before_attrition = ceiling(n_single_mean),
  final_sample_size = n_single_mean_final
)

kable(single_mean_result, digits = 3, caption = "Sample size for estimating a single mean")
```

# 9) Scenario 3: two independent proportions

For equal allocation and a two-sided superiority test:

\[n_{per\ group} = \frac{\left[Z_{1-\alpha/2}\sqrt{2\bar{p}(1-\bar{p})} +
Z_{1-\beta}\sqrt{p_1(1-p_1) + p_2(1-p_2)}\right]^2}{(p_1 - p_2)^2}
\]

```{r two-proportions}
p1 <- risk_control
p2 <- risk_intervention
p_bar <- (p1 + p2) / 2
risk_difference <- abs(p1 - p2)

n_two_prop_group <- (
  z_alpha * sqrt(2 * p_bar * (1 - p_bar)) +
    z_power * sqrt(p1 * (1 - p1) + p2 * (1 - p2))
)^2 / risk_difference^2

n_two_prop_group_final <- ceiling(n_two_prop_group / (1 - attrition))

# Optional alternative using base R:
# power.prop.test(p1 = p1, p2 = p2, power = power, sig.level = alpha)$n

two_prop_result <- tibble(
  scenario = "Two independent proportions",
  control_event_risk = p1,
  intervention_event_risk = p2,
  absolute_risk_difference = p2 - p1,
  sample_size_per_group_before_attrition = ceiling(n_two_prop_group),
  final_sample_size_per_group = n_two_prop_group_final,
  final_total_sample_size = 2 * n_two_prop_group_final
)

kable(two_prop_result, digits = 3, caption = "Sample size for comparing two proportions")
```

# 10) Scenario 4: two independent means

\[n_{per\ group} = \frac{2\sigma^2\left(Z_{1-\alpha/2} + Z_{1-\beta}\right)^2}{\Delta^2}\]

```{r two-means}
sigma <- continuous_sd
delta <- minimal_difference_mean

n_two_means_group <- 2 * sigma^2 * (z_alpha + z_power)^2 / delta^2
n_two_means_group_final <- ceiling(n_two_means_group / (1 - attrition))

# Optional alternative using base R:
# power.t.test(delta = delta, sd = sigma, power = power,
#              sig.level = alpha, type = "two.sample")$n

two_means_result <- tibble(
  scenario = "Two independent means",
  expected_sd = sigma,
  clinically_meaningful_difference = delta,
  sample_size_per_group_before_attrition = ceiling(n_two_means_group),
  final_sample_size_per_group = n_two_means_group_final,
  final_total_sample_size = 2 * n_two_means_group_final
)

kable(two_means_result, digits = 3, caption = "Sample size for comparing two means")
```

# 11) Scenario 5: unmatched case-control study

The expected exposure among cases can be approximated from the exposure prevalence among controls and the odds ratio:

\[
p_1 = \frac{OR \times p_0}{1 - p_0 + OR \times p_0}
\]

```{r case-control}
p0 <- p_exposed_controls
p1 <- odds_ratio * p0 / (1 - p0 + odds_ratio * p0)
p_bar <- (p0 + p1) / 2
exposure_difference <- abs(p1 - p0)

n_case_control_group <- (
  z_alpha * sqrt(2 * p_bar * (1 - p_bar)) +
    z_power * sqrt(p0 * (1 - p0) + p1 * (1 - p1))
)^2 / exposure_difference^2

n_case_control_group_final <- ceiling(n_case_control_group / (1 - attrition))

case_control_result <- tibble(
  scenario = "Unmatched case-control study",
  exposure_prevalence_controls = p0,
  expected_odds_ratio = odds_ratio,
  exposure_prevalence_cases = p1,
  sample_size_per_group_before_attrition = ceiling(n_case_control_group),
  final_sample_size_per_group = n_case_control_group_final,
  final_total_sample_size = 2 * n_case_control_group_final
)

kable(case_control_result, digits = 3, caption = "Sample size for an unmatched case-control study")
```

# 12) Scenario 6: diagnostic-accuracy study

For sensitivity:

\[n_{diseased} = \frac{Z_{1-\alpha/2}^{2}Se(1-Se)}{d_{Se}^{2}}\]

For specificity:

\[n_{non-diseased} = \frac{Z_{1-\alpha/2}^{2}Sp(1-Sp)}{d_{Sp}^{2}}\]

```{r diagnostic-accuracy}
n_diseased <- z_alpha^2 * sensitivity * (1 - sensitivity) / precision_sensitivity^2
n_nondiseased <- z_alpha^2 * specificity * (1 - specificity) / precision_specificity^2

total_for_sensitivity <- ceiling(n_diseased / disease_prevalence)
total_for_specificity <- ceiling(n_nondiseased / (1 - disease_prevalence))
total_diagnostic <- max(total_for_sensitivity, total_for_specificity)
total_diagnostic_final <- ceiling(total_diagnostic / (1 - attrition))

diagnostic_result <- tibble(
  scenario = "Diagnostic-accuracy study",
  expected_sensitivity = sensitivity,
  expected_specificity = specificity,
  disease_prevalence = disease_prevalence,
  diseased_required = ceiling(n_diseased),
  nondiseased_required = ceiling(n_nondiseased),
  total_before_attrition = total_diagnostic,
  final_total_sample_size = total_diagnostic_final
)

kable(diagnostic_result, digits = 3, caption = "Sample size for a diagnostic-accuracy study")
```

# 13) Final summary

```{r final-summary}
final_summary_table <- tibble(
  Scenario = c(
    "Prevalence survey",
    "Single mean",
    "Two independent proportions",
    "Two independent means",
    "Unmatched case-control study",
    "Diagnostic-accuracy study"
  ),
  Sample_size = c(
    prevalence_result$final_sample_size,
    single_mean_result$final_sample_size,
    two_prop_result$final_total_sample_size,
    two_means_result$final_total_sample_size,
    case_control_result$final_total_sample_size,
    diagnostic_result$final_total_sample_size
  ),
  Note = c(
    "Total participants",
    "Total participants",
    "Total participants",
    "Total participants",
    "Cases plus controls",
    "Total participants"
  )
)

kable(final_summary_table, digits = 0, caption = "Final summary table of sample-size calculations")
```

# End of document
