---
title: "Relative and absolute measures of effect in clinical research: interpretation of risk metrics and their ratios with computational implementation in r and python"
authors: "Renato Carneiro de Freitas Chaves, Tiago Mendonça dos Santos, Thiago Domingos Corrêa"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# 1) Purpose

This R Markdown document provides a step-by-step structure to calculate effect measures for a **binary outcome** comparing:
- **Control group** (`group = 0`)
- **Intervention group** (`group = 1`)

Effect measures:

- **Relative Risk (RR)**
- **Relative Risk Reduction (RRR)**
- **Absolute Risk Reduction (ARR)**
- **Number Needed to Treat (NNT)**
- **Number Needed to Harm (NNH)**

## Outcome definition in this dataset

In the provided dataset model (`data.xlsx`), the binary endpoint is:
- `outcome`: `0 = Alive`, `1 = Dead`
We treat **death** (`outcome = 1`) as the **event of interest**.

> **Interpretation rule**  
> - If the intervention **reduces** event risk, report **NNT**.  
> - If the intervention **increases** event risk, report **NNH**.

# 2) Packages

```{r packages}
# Install once if necessary:
# install.packages(c("readxl", "dplyr", "janitor", "epitools", "tibble", "skimr"))

library(readxl)
library(dplyr)
library(janitor)
library(epitools)
library(tibble)
library(skimr)
```

# 3) Data Import (Excel)
This guide uses the Excel file **data.xlsx** with a sheet named **"Data"**.

- If `data.xlsx` is in the same folder as this `.Rmd`, you can keep the path as `"data.xlsx"`.
- Otherwise, replace with the full path to your file.


```{r import-data}
data <- read_excel("data.xlsx", sheet = "Data") 

```

# 4) Confirm Required Variables

For the requested calculations, we need:
- `group` (0 = Control, 1 = Intervention)
- `outcome` (0 = Alive, 1 = Dead)

```{r verify-vars-2}
# Quick frequency checks
skimr::skim(data)

data %>% 
  tabyl(outcome)
```
# 5) Create the Analysis Dataset

We will:

- keep only rows with non-missing `group` **and** `outcome`
- ensure they are coded as expected (0/1)
- define the event as `outcome == 1` (death)

```{r analysis-dataset}
analysis_data <- data %>%
  select(group, outcome) %>%
  filter(!is.na(group), !is.na(outcome)) %>%
  mutate(
    group = as.integer(group),
    event = as.integer(outcome)  # event = 1 means "death"
  )

```
# 6) Construct the 2×2 Table

Rows:
- Intervention
- Control


Columns:
- Event (Dead)
- Non-event (Alive)

```{r table-2x2}
analysis_data <- analysis_data %>%
  mutate(
    group_label = factor(group, levels = c(1, 0), labels = c("Intervention", "Control")),
    event_label = factor(event, levels = c(1, 0), labels = c("Event", "Non-event"))
  )

tab <- table(analysis_data$group_label, analysis_data$event_label)
tab
```

# 7) Compute Risks and Effect Measures

## 7.1 Extract cell counts (a, b, c, d)

We define:

- `a` = Events in **Intervention**
- `b` = Non-events in **Intervention**
- `c` = Events in **Control**
- `d` = Non-events in **Control**

```{r counts}
a <- tab[["Intervention", "Event"]]
b <- tab[["Intervention", "Non-event"]]
c <- tab[["Control", "Event"]]
d <- tab[["Control", "Non-event"]]
```

## 7.2 Compute risks (event rates)

\[ Risk_{Intervention} = \frac{a}{a+b}, \quad Risk_{Control} = \frac{c}{c+d} \]

```{r risks}
(risk_intervention <- a / (a + b))
(risk_control <- c / (c + d))
```

## 7.3 Relative Risk (RR)

\[ RR = \frac{Risk_{Intervention}}{Risk_{Control}} \]

```{r rr}
RR <- risk_intervention / risk_control
RR
```

## 7.4 Relative Risk Reduction (RRR)

When the event is undesirable (death), a definition is:

\[ RRR = 1 - RR \]

```{r rrr}
RRR <- 1 - RR
RRR
```

## 7.5 Absolute Risk Reduction (ARR)

\[ ARR = Risk_{Control} - Risk_{Intervention} \]

```{r arr}
ARR <-  risk_control - risk_intervention
ARR
```

## 7.6 NNT and NNH

- If `ARR > 0`, the intervention reduces risk → **NNT = 1 / ARR**
- If `ARR < 0`, the intervention increases risk → **NNH = 1 / |ARR|**
- If `ARR = 0`, neither NNT nor NNH is defined (infinite)

Since the absolute risk reduction is positive ($ARR =$ `r round(ARR, 2)`), the number needed to treat (NNT = 1/ARR) is `r round(1/ARR, 2)`.

```{r nnt-nnh,  echo = FALSE}
if (ARR > 0) {
  NNT <- 1 / ARR
  NNH <- NA_real_
} else if (ARR < 0) {
  NNT <- NA_real_
  NNH <- 1 / abs(ARR)
} else {
  NNT <- Inf
  NNH <- Inf
}

# NNT
# NNH
```

# 8) Confidence Intervals for Risk Difference and Relative Risk

For reporting, the **Absolute risk reduction (ARR)** and **relative risk (RR)** should be presented with **95% confidence intervals (CIs)**.

- **Absolute risk reduction(ARR)** is reported as:  
  \[ ARR = Risk_{Control} - Risk_{Intervention} \]
- **Relative risk (RR)** is reported as:  
  \[ RR = \frac{Risk_{Intervention}}{Risk_{Control}} \]

## 8.1 Absolute risk reduction (ARR) with 95% CI
A Wald-type CI for the difference in proportions is obtained using `prop.test(..., correct = FALSE)`.

```{r rd-ci}
arr_test <- prop.test(
  x = c(a, c),         # events: Intervention, Control
  n = c(a + b, c + d), # totals: Intervention, Control
  correct = FALSE
)

ARR <- arr_test$estimate[[2]] - arr_test$estimate[[1]]  # intervention - control
ARR_CI_low  <- arr_test$conf.int[2]  
ARR_CI_high <- arr_test$conf.int[1]
```

## 8.2 Relative risk (RR) with 95% CI

A log (Wald) CI is computed for the RR. To ensure stability when any cell count is zero, a continuity correction of 0.5 is applied.

```{r rr-ci}
# Continuity correction if any cell is zero
cc <- ifelse(any(c(a, b, c, d) == 0), 0.5, 0)

a_cc <- a + cc
b_cc <- b + cc
c_cc <- c + cc
d_cc <- d + cc

risk_intervention_cc <- a_cc / (a_cc + b_cc)
risk_control_cc      <- c_cc / (c_cc + d_cc)

RR_cc <- risk_intervention_cc / risk_control_cc

SE_logRR <- sqrt((1 / a_cc) - (1 / (a_cc + b_cc)) + (1 / c_cc) - (1 / (c_cc + d_cc)))

alpha <- 0.05
z <- qnorm(1 - alpha / 2)

RR_CI_low  <- exp(log(RR_cc) - z * SE_logRR)
RR_CI_high <- exp(log(RR_cc) + z * SE_logRR)

tibble(
  rr = RR_cc,
  rr_ci_low = RR_CI_low,
  rr_ci_high = RR_CI_high,
  continuity_correction_applied = cc
)
```

# 9) Final Summary Table

```{r summary-table}
# A compact, publication table with CI columns where applicable
final_summary_table <- tribble(~measure,    ~value,            ~lower_ci,  ~upper_ci, 
"Risk (Control)",                            risk_control,      NA,         NA,
"Risk (Intervention)",                       risk_intervention, NA,         NA,
"Absolute Risk Reduction (ARR = Rc - Ri)",   ARR,               ARR_CI_low, ARR_CI_high,
"Relative Risk (RR)",                        RR,                RR_CI_low,  RR_CI_high,
"Relative Risk Reduction (RRR = 1 - RR)",    RRR,               NA,         NA,
"Number Needed to Treat (NNT) [if ARR > 0]", NNT,               NA,         NA, 
"Number Needed to Harm (NNH) [if ARR < 0]",  NNH,               NA,         NA
)

final_summary_table
```

---

## End of document
